P3064R2: How to Avoid OOTA Without Really Trying

### 2024-08-22 paper/ootavacuous.tex
ootavacuous: Remove "User Influence Over Language Semantics" appendix
Instead, tell the reader how to get this information from the Linux
kernel.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

The exact definition of a computer language is a subject of some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each being prone to change over time.
In areas that are not well settled or where users might reasonably
want to resist the dictates of the standard,
compilers often provide switches to override their default behaviors.
An example is GCC's \co{-funsigned-char} command-line argument,
which causes it to treat variables of type \co{char} as unsigned.
More examples of user control over language semantics are given in
Appendix~\ref{app:User Influence Over Language Semantics}
and by the discussion
in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.

@@

\section{User Influence Over Language Semantics}
\label{app:User Influence Over Language Semantics}

As noted earlier, the exact definition of a computer language is subject
to some debate, with standards, implementations, and users all having
some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
It is natural to dismiss user influence when compared to the text of
standards or the code in implementations, but both of these are subject to
change and do change over time.
An especially easy way for users to influence the implementation is by
means of command-line flags and switch settings.
As an example,
consider the following command line used to compile the Linux kernel's
\co{kernel/rcu/tree.c} C-language source file:

\begin{quote}
{
	\scriptsize
	\texttt{\noindent
	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
	-I./arch/x86/include \\
	-I./arch/x86/include/generated  -I./include
	-I./arch/x86/include/uapi \\
	-I./arch/x86/include/generated/uapi
	-I./include/uapi \\
	-I./include/generated/uapi \\
	-include ./include/linux/compiler-version.h \\
	-include ./include/linux/kconfig.h \\
	-include ./include/linux/compiler\_types.h \\
	-D\_\_KERNEL\_\_
	-fmacro-prefix-map=./=
	-Werror -std=gnu11 -fshort-wchar \\
	-funsigned-char -fno-common -fno-PIE
	-fno-strict-aliasing \\
	-mno-sse
	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx \\
	-fcf-protection=branch
	-fno-jump-tables -m64 -falign-jumps=1 \\
	-falign-loops=1 -mno-80387 -mno-fp-ret-in-387 \\
	-mpreferred-stack-boundary=3 -mskip-rax-setup \\
	-mtune=generic -mno-red-zone -mcmodel=kernel
	-Wno-sign-compare \\
	-fno-asynchronous-unwind-tables
	-mindirect-branch=thunk-extern \\
	-mindirect-branch-register
	-mindirect-branch-cs-prefix \\
	-mfunction-return=thunk-extern -fno-jump-tables \\
	-fpatchable-function-entry=16,16
	-fno-delete-null-pointer-checks \\
	-O2 -fno-allow-store-data-races
	-fstack-protector-strong \\
	-fomit-frame-pointer
	-fno-stack-clash-protection -falign-functions=16 \\
	-fno-strict-overflow -fno-stack-check -fconserve-stack
	-Wall -Wundef \\
	-Werror=implicit-function-declaration
	-Werror=implicit-int \\
	-Werror=return-type
	-Werror=strict-prototypes -Wno-format-security \\
	-Wno-trigraphs -Wno-frame-address \\
	-Wno-address-of-packed-member
	-Wframe-larger-than=2048 -Wno-main \\
	-Wno-unused-but-set-variable -Wno-unused-const-variable -Wvla \\
	-Wno-pointer-sign -Wcast-function-type -Wno-array-bounds \\
	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5 \\
	-Werror=date-time -Werror=incompatible-pointer-types \\
	-Werror=designated-init -Wenum-conversion
	-Wno-unused-but-set-variable \\
	-Wno-unused-const-variable -Wno-restrict -Wno-packed-not-aligned \\
	-Wno-format-overflow 
	-Wno-format-truncation -Wno-stringop-overflow \\
	-Wno-stringop-truncation
	-Wno-missing-field-initializers \\
	-Wno-type-limits
	-Wno-shift-negative-value -Wno-maybe-uninitialized \\
	-Wno-sign-compare 
	-DKBUILD\_MODFILE='"kernel/rcu/tree"' \\
	-DKBUILD\_BASENAME='"tree"' -DKBUILD\_MODNAME='"tree"' \\
	-D\_\_KBUILD\_MODNAME=kmod\_tree \\
	-c -o kernel/rcu/tree.o kernel/rcu/tree.c
	}
}
\end{quote}

We do not propose to explain all of these, and sufficiently motivated
readers can avail themselves of the GCC documentation.
We instead look at representative members of several categories.

The \co{-funsigned-char} causes the \co{char} type to be unsigned, which
overrides per-architecture defaults, some of which treat \co{char}
as signed and others as unsigned.
This choice prevents a class of bugs, and also allows the kernel to make
reliable use of the uppermost bit of variables of type \co{char}.
It also affects the definition of ``semantic dependency'' by changing
the arithmetic proporties of this type.
In theory, the standard could have specified the signedness of \co{char}
but the variety of existing practice in the 1980s prevented this.
Plus, optimizers were much less capable back then, so the signedness of
\co{char} was much more of a performance concern than it is today.

The \co{-mno-sse} prevents GCC from making use of the processor's SSE hardware.
This is done for performance reasons, as it avoids the overhead of
saving and restoring the state of this hardware when switching between
user and kernel contexts.
Similarly, the \co{-mcmodel=kernel} causes the kernel binary to be
placed in the uppermost 2GB of the address space, again reducing
the overhead of switching between user and kernel contexts.
These are cases where the standard does not specify anything, nor
should it.

The \co{-fpatchable-function-entry=16,16} causes GCC to emit 16
\co{nop} instructions at the beginning of each function, with
the function's entry point being just after this string of \co{nop}
instructions.
The resulting buffers are used by the Linux-kernel tracing infrastructure
which is in turn used for debugging, performance measurement, and
monitoring.
This is again clearly outside the scope of the standard.

The \co{-fstack-protector-strong} causes GCC to emit code that provides
some protection against some classes of attacks based on buffer overflows.
One could rightly argue code should simply avoid ever overflowing buffers,
but things like memory allocators and userspace memory accesses must
use code that can be difficult to distinguish from buffer overflows.
It is not clear that the ever-increasing variety of attacks should
affect the standard.

The \co{-fno-strict-overflow} causes GCC to act as if signed integer
overflow is defined behavior,
which also affects the definition of ``semantic dependency''.
This might be a controversial choice, and another option would be to add
a new set of signed integer types to the standard for which overflow is
defined as wrapping, similar to the situation with unsigned integers.

The \co{-Werror=strict-prototypes} causes GCC to warn if old-style
non-ANSI function prototypes are used.
This helps avoid certain classes of bugs.
Warnings are by design outside the scope of the standard.

To sum up, user preference can exert a nontrivial influence over
language semantics, and in particular can affect some aspects of the
definition of ``semantic dependency''.

>>>>

The exact definition of a computer language is a subject of some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each being prone to change over time.
In areas that are not well settled or where users might reasonably
want to resist the dictates of the standard,
compilers often provide switches to override their default behaviors.
An example is GCC's \co{-funsigned-char} command-line argument,
which causes it to treat variables of type \co{char} as unsigned.
Many more examples of user control over language semantics
may be found through use of the command \co{make V=1}
in a Linux-kernel source tree and by the discussion
in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.

@@
